---
title: "Tutorial #2: Structuring a Data Story for Policy Audiences"
subtitle: "How to sequence your findings so policymakers reach the right conclusion before you say a word"
author: "Olamide Akeboi"
date: "2026-04-02"
order: 2
categories: [data visualisation, R, policy, communication, beginners]
toc: true
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## The problem

Your analysis is solid. Your charts are clean. You walk into the room confident.

And then the policymaker asks: "So what does this mean for us?"

You have just spent thirty minutes presenting. They should know what it means. But they are asking because your presentation answered the wrong questions in the wrong order — and left the most important question unanswered entirely.

This tutorial is about fixing that. Not the chart. The structure around it.

------------------------------------------------------------------------

## What you will learn

By the end of this tutorial you will be able to:

-   Explain why most data presentations fail to drive decisions
-   Apply the What-So What-Now What framework to structure any policy presentation
-   Use the Pyramid Principle to lead with your conclusion, not your methodology
-   Humanise data so numbers connect to real decisions about real people
-   Build an annotated grouped bar chart in R that highlights two distinct policy problems

------------------------------------------------------------------------

## Prerequisites

Before starting, make sure you have completed Tutorial #1 or have the following packages installed:

```{r}
#| eval: false
install.packages(c("ggplot2", "ggtext", "patchwork", "scales", "dplyr", "tidyr"))
```

------------------------------------------------------------------------

## Why structure matters more than data

Policymakers are not passive recipients of information. They are active filters — constantly scanning for what is relevant, what is urgent, and what requires a decision from them specifically.

Research on how policymakers engage with evidence shows three consistent patterns:

**They are time-poor.** A policymaker reviewing fifteen slides in a thirty-minute meeting has roughly two minutes per slide. They will not read every word. They will scan for the headline and move on.

**They are not subject-matter experts.** Handing a policymaker raw data or complex statistics without interpretation does not inform them — it burdens them. The interpretation is your job, not theirs.

**They filter through a specific set of questions.** Whether they articulate it or not, every policymaker sitting in a data presentation is asking three things: what are the facts, why does this matter to me, and what should I do about it?

Most analytical presentations answer the first question thoroughly, gesture vaguely at the second, and leave the third entirely to the audience.

That is not a data problem. It is a structure problem.

------------------------------------------------------------------------

## Data

We will use zonal Penta1 and Penta3 coverage data from the **2023–24 Nigeria Demographic and Health Survey (DHS)**, available at [dhsprogram.com](https://dhsprogram.com/pubs/pdf/PR157/PR157.pdf).

Penta1 is the first dose of the pentavalent vaccine. Penta3 is the third and final dose. The gap between them — the dropout — tells us how many children started vaccination but never completed it. This is one of the most actionable insights in immunisation programming because these children are not unreached — they are under-served.

```{r}
# Load packages
library(ggplot2)
library(ggtext)
library(patchwork)
library(scales)
library(dplyr)
library(tidyr)

# Zonal Penta1 and Penta3 coverage from DHS 2023-24 (PR157)
zone_data <- data.frame(
  zone   = c(
    "North Central", "North East", "North West",
    "South East", "South South", "South West"
  ),
  penta1 = c(60.4, 64.3, 46.1, 93.1, 87.2, 84.4),
  penta3 = c(48.2, 53.0, 39.7, 79.3, 79.6, 72.8)
) %>%
  mutate(
    dropout     = penta1 - penta3,
    dropout_pct = round((dropout / penta1) * 100, 1)
  )
```

`mutate()` creates new columns. Here we calculate two things — the absolute dropout (the gap in percentage points between Penta1 and Penta3) and the dropout rate (the proportion of children who started but did not finish, expressed as a percentage of those who started).

------------------------------------------------------------------------

## Inspect the data

Before building any chart, look at what the data is telling you.

```{r}
zone_data
```

Take a moment to read the dropout column carefully. Two patterns should immediately stand out:

The North West has the lowest Penta1 coverage nationally — meaning children are not even entering the vaccination system. This is an **access problem**.

The South East has the highest absolute dropout despite the strongest Penta1 coverage — meaning children are starting but not finishing. This is a **continuity problem**.

These are two different policy problems requiring two different interventions. A chart that shows only Penta3 would hide this distinction entirely.

------------------------------------------------------------------------

## Framework 1 — What? So What? Now What?

This is the simplest and most powerful framework for structuring any policy presentation. It maps directly onto the three questions every policymaker is asking.

**What?** — Clearly describe the facts. This is the evidence layer.

**So What?** — Explain the implications. Why does this matter? This is the interpretation layer.

**Now What?** — Provide a clear, actionable recommendation. This is the decision layer.

Let us build a chart for each layer using the same data.

### What? — The evidence

```{r}
#| fig-height: 5

# Reshape to long format for grouped bar chart
zone_long <- zone_data %>%
  pivot_longer(
    cols      = c(penta1, penta3),
    names_to  = "dose",
    values_to = "coverage"
  ) %>%
  mutate(
    dose = case_when(
      dose == "penta1" ~ "Penta1 (first dose)",
      dose == "penta3" ~ "Penta3 (third dose)"
    )
  )

# Order zones by Penta1 coverage descending
zone_order <- zone_data %>%
  arrange(desc(penta1)) %>%
  pull(zone)

zone_long <- zone_long %>%
  mutate(zone = factor(zone, levels = zone_order))

# What chart — shows the pattern without interpretation
chart_what <- ggplot(
  zone_long,
  aes(
    x    = zone,
    y    = coverage,
    fill = dose
  )
) +
  geom_col(
    position = position_dodge(width = 0),
    width    = 0.7
  ) +
  scale_fill_manual(
    values = c(
      "Penta1 (first dose)" = "#534AB7",
      "Penta3 (third dose)" = "#A89EDB"
    )
  ) +
  scale_y_continuous(
    labels = label_percent(scale = 1),
    limits = c(0, 100)
  ) +
  labs(
    title   = "Penta1 and Penta3 coverage by zone, Nigeria 2023–24",
    x       = NULL,
    y       = "Coverage (%)",
    fill    = NULL,
    caption = "Source: Nigeria DHS 2023–24 (PR157)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "top",
    axis.text.x     = element_text(angle = 15, hjust = 1)
  )

chart_what
```

This chart answers What. It shows Penta1 and Penta3 coverage side by side for each zone. But it leaves the reader to figure out the implication — and labels nothing about what it means for decisions.

### So What? — The interpretation

Now we rebuild the same chart to answer So What. We add three things: background highlights on the two zones with distinct problems, dropout gap annotations, and problem labels in the legend so the reader understands the argument before studying a single bar.

```{r}
#| fig-height: 6

# Set zone factor order
zone_data <- zone_data %>%
  mutate(zone = factor(zone, levels = zone_order))

zone_long <- zone_long %>%
  mutate(zone = factor(zone, levels = zone_order))

# Add dropout gap to zone_long for annotations
zone_long_annotated <- zone_long %>%
  left_join(
    zone_data %>% select(zone, dropout, penta1),
    by = "zone"
  )

# So What chart
chart_sowhat <- ggplot(
  zone_long,
  aes(
    x    = zone,
    y    = coverage,
    fill = dose
  )
) +
  # Background highlight — South East (continuity problem)
  annotate(
    geom  = "rect",
    xmin  = 0.5, xmax = 1.5,
    ymin  = 0,   ymax = 100,
    fill  = "#D85A30",
    alpha = 0.08
  ) +
  # Background highlight — North West (access problem)
  annotate(
    geom  = "rect",
    xmin  = 5.5, xmax = 6.5,
    ymin  = 0,   ymax = 100,
    fill  = "#534AB7",
    alpha = 0.08
  ) +
  # Bars — touching within each zone
  geom_col(
    position = position_dodge(width = 0.75),
    width    = 0.75
  ) +
  # Dropout gap segments
  geom_segment(
    data = zone_data,
    aes(
      x     = as.numeric(zone) + 0.19,
      xend  = as.numeric(zone) + 0.19,
      y     = penta3,
      yend  = penta1,
      group = zone
    ),
    colour    = "#AAAAAA",
    linewidth = 0.5,
    inherit.aes = FALSE
  ) +
  # Dropout gap labels
  geom_text(
    data = zone_data,
    aes(
      x     = as.numeric(zone) + 0.32,
      y     = penta3 + (dropout / 2),
      label = paste0(round(dropout, 1), "pp")
    ),
    size        = 2.8,
    colour      = "#888888",
    inherit.aes = FALSE
  ) +
  scale_fill_manual(
    values = c(
      "Penta1 (first dose)"                               = "#534AB7",
      "Penta3 (third dose)"                               = "#A89EDB"
    )
  ) +
  scale_y_continuous(
    labels = label_percent(scale = 1),
    limits = c(0, 105)
  ) +
  labs(
    title    = "**Every zone loses children between first and third dose**",
    subtitle = "The North West fails at entry. The South East fails at completion. Both need urgent attention.",
    x        = NULL,
    y        = "Coverage (%)",
    fill     = NULL,
    caption  = "Source: Nigeria DHS 2023–24 (PR157)\nNote: A small dropout gap in the North West does not indicate strong performance — absolute coverage remains the lowest nationally."
  ) +
  # Problem labels as annotations inside the highlighted zones
  annotate(
    geom     = "text",
    x        = 1,
    y        = 103,
    label    = "Continuity problem",
    colour   = "#D85A30",
    size     = 3,
    fontface = "bold"
  ) +
  annotate(
    geom     = "text",
    x        = 6,
    y        = 103,
    label    = "Access problem",
    colour   = "#534AB7",
    size     = 3,
    fontface = "bold"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position = "top",
    axis.text.x     = element_text(angle = 15, hjust = 1),
    plot.title      = element_markdown(size = 12, lineheight = 1.3),
    plot.subtitle   = element_text(size = 10, colour = "#555555"),
    plot.caption    = element_text(size = 7.5, colour = "#999999")
  )

chart_sowhat
```

::: callout-note
A common misreading of this chart: the North West appears to have a small dropout gap (6.4pp) compared to the South East (13.8pp). This does not mean the North West is performing better. The North West starts at 46.1% Penta1 — the lowest entry point in the country. The access problem is so severe that few children enter the system in the first place. The South East starts at 93.1% but loses 13.8 percentage points before completion. These are structurally different problems requiring structurally different responses.
:::

### Now What? — The recommendation

The Now What belongs in your spoken narrative, executive summary, or policy brief. Here is what it looks like for this data:

> *"The North West has a clear access problem — the lowest Penta1 coverage nationally at 46.1% means children are not entering the immunisation system. The priority intervention here is outreach, demand generation, and community engagement to bring first-time recipients into the system. The South East has a clear continuity problem — strong entry at 93.1% Penta1 but the highest dropout nationally at 13.8 percentage points means children are starting but not finishing. The priority intervention here is defaulter tracing, reminder systems, and caregiver follow-up. The remaining zones — North East, North Central, and South West — present mixed challenges with both moderate access gaps and significant dropout. These require zone-specific diagnostic work before a single intervention type is prescribed. A uniform national strategy will not address these structurally different problems equally."*

Notice what this recommendation does:

-   It names each geography and its specific problem separately
-   It quantifies the problem — "46.1% Penta1", "13.8 percentage points"
-   It names the mechanism for each — access vs continuity interventions
-   It acknowledges complexity in the middle zones rather than oversimplifying
-   It closes with a structural argument — one strategy will not work for all zones

------------------------------------------------------------------------

## Framework 2 — The Pyramid Principle

The Pyramid Principle flips the traditional academic structure. Instead of building toward your conclusion through evidence, you lead with your conclusion and support it beneath.

**Traditional academic structure:** Context → Data → Analysis → Finding → Recommendation

**Pyramid structure:** Recommendation → Key findings → Supporting evidence

Applied to our data:

**Top — the recommendation:** "Design zone-specific interventions — access strategies for the north, continuity strategies for the south."

**Middle — key findings:**

-   Every zone loses children between Penta1 and Penta3
-   The North West starts lowest at 46.1% Penta1 — an access problem
-   The South East has the largest absolute dropout at 13.8 percentage points — a continuity problem
-   No northern zone reaches the 80% Penta3 target

**Base — supporting evidence:** The full zonal breakdown, state-level data, trend analysis over time.

For a policymaker with two minutes per slide, only the top of the pyramid may get read. The Pyramid Principle ensures that if they read only one thing, they read the most important thing.

------------------------------------------------------------------------

## Framework 3 — Humanising the data

Numbers alone do not motivate action. Stories do.

Anchor your data with a human reality — name who is affected, what the consequence is, and what is preventable.

```{r}
# Calculate humanised framings for each zone
zone_data %>%
  mutate(
    unvaccinated_per_100 = round(100 - penta3),
    dropout_per_100      = round(dropout),
    humanised = paste0(
      zone, ": ", dropout_per_100,
      " in every 100 children who started vaccination never completed it"
    )
  ) %>%
  select(zone, penta1, penta3, dropout_per_100, humanised)
```

Compare these two framings of the South East finding:

**Data only:** "The South East has a Penta1 to Penta3 dropout of 13.8 percentage points."

**Humanised:** "In the South East — Nigeria's highest-performing region — nearly 1 in 7 children who received their first vaccine dose never completed the series. These children are not unreached. They are under-served. And because they are already in the system, they are findable."

The second framing names the children, names the paradox, and names the opportunity. It gives the policymaker something to act on behalf of — not a percentage point, but a child who started and was lost.

------------------------------------------------------------------------

## Putting it all together

```{r}
#| fig-height: 6
#| fig-width: 13

chart_what + chart_sowhat +
  plot_annotation(
    title = "Same data. Structured differently. Different decisions.",
    theme = theme(
      plot.title = element_text(size = 14, face = "bold", hjust = 0.5)
    )
  ) +
  plot_layout(widths = c(1, 1.2))
```

The left panel shows the data. The right panel makes the argument. The background highlights, gap annotations, and problem labels do the interpretive work so the policymaker does not have to.

------------------------------------------------------------------------

## A checklist before you present

Before your next policy presentation run through these five questions:

**1. Have I answered What, So What, and Now What?** If any layer is missing your presentation is incomplete.

**2. Does my opening statement lead with the conclusion?** If you are building toward your finding you have lost half the room before you get there.

**3. Have I distinguished between different types of problems in the data?** The same indicator can reflect different problems in different geographies. Name them separately.

**4. Is my recommendation specific enough to act on?** "Strengthen immunisation systems" is not a recommendation. "Design defaulter tracing systems in South East zones with Penta1 above 80% and dropout above 10%" is.

**5. If they read only the title of each chart — have I said enough?** Return to Tutorial #1 if the answer is no.

------------------------------------------------------------------------

## Your turn

Using your own data, apply the What-So What-Now What framework:

-   Build a What chart — descriptive, shows the pattern
-   Build a So What chart — active title, names the implication
-   Write the Now What in one paragraph — specific, actionable, time-bound

Share your before and after in the comments.

------------------------------------------------------------------------

## Further resources

-   [Storytelling with Data — Cole Nussbaumer Knaflic](https://www.storytellingwithdata.com/)
-   [2023–24 Nigeria DHS Key Indicators Report](https://dhsprogram.com/pubs/pdf/PR157/PR157.pdf)

------------------------------------------------------------------------

## Next tutorial

**Tutorial #3: Bar, line, map, or table? Choosing the right chart type for policy data in R**

*Publishing Wednesday 9 April 2026.*
