---
title: "Tutorial #3: Choosing the Right Chart Type for Policy Data"
subtitle: "How to match your visual to the decision your audience needs to make"
author: "Olamide Akeboi"
date: "2026-05-14"
order: 3
categories: [data visualisation, R, policy, beginners]
toc: true
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## What this tutorial covers

Most chart mistakes in public health are not about aesthetics. They are about mismatch.

A pie chart used to show change over time. A line chart used to compare categories with no natural order. A bar chart stacked so heavily that no individual segment can be read. In each case, the analyst made a chart. They just made the wrong one.

The chart type you choose determines what comparison your audience can make. Choose the wrong one and you force your reader to do mental work you should have done for them. In a policy meeting, that mental work rarely gets done. The chart gets skipped.

This tutorial gives you a practical decision framework for matching chart type to comparison type, with worked examples using Nigeria DHS 2023--24 data throughout.

By the end, you will be able to:

-   Identify the comparison type behind any data question
-   Select the appropriate chart type for that comparison
-   Build five core chart types in R using `ggplot2`
-   Recognise the most common chart type mismatches in public health data

------------------------------------------------------------------------

## The core principle: charts encode comparisons

Every chart makes a comparison. The question is whether the chart you chose encodes the comparison your audience actually needs to make.

There are five comparison types that cover the vast majority of public health data questions:

| Comparison type | The question you are answering |
|---|---|
| **Ranking** | Which states, zones, or groups performed best or worst? |
| **Part-to-whole** | What share of the total does each group represent? |
| **Change over time** | How has this indicator moved across time? |
| **Distribution** | How spread out are the values? Where are the extremes? |
| **Relationship** | Does one variable move with another? |

Each comparison type has chart types that work well for it and chart types that actively obstruct it. The sections below work through each one.

------------------------------------------------------------------------

## Setup

```{r}
#| label: setup

library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(forcats)
```

------------------------------------------------------------------------

## Comparison type 1: Ranking

**The question:** Which states, zones, or groups performed best or worst?

**Chart type to use:** Horizontal bar chart, ordered by value.

**Why not a vertical bar chart?** Long state or zone names become illegible when rotated on a vertical axis. Horizontal bars give your labels room to breathe. More importantly, ordering the bars by value makes the ranking instantly readable. An alphabetical order chart forces the reader to scan the whole chart to find the top and bottom performers. An ordered chart tells the story in the first second.

### Example: Penta3 coverage by state, Nigeria DHS 2023--24

The data below shows estimated Penta3 coverage (third dose of the pentavalent vaccine) by state from the 2023--24 Nigeria DHS PR157 report.

```{r}
#| label: ranking-data

penta3_state <- data.frame(
  state = c(
    "Anambra", "Lagos", "Ekiti", "Imo", "Osun", "Ogun", "Abia",
    "Rivers", "Oyo", "Enugu", "Edo", "Ondo", "Cross River", "Delta",
    "Ebonyi", "Akwa Ibom", "Benue", "Nasarawa", "Plateau", "Taraba",
    "FCT Abuja", "Niger", "Kwara", "Kogi", "Kaduna", "Sokoto",
    "Katsina", "Kano", "Jigawa", "Zamfara", "Yobe", "Borno",
    "Bauchi", "Gombe", "Adamawa", "Kebbi"
  ),
  coverage = c(
    95.4, 91.6, 90.3, 89.8, 89.5, 88.2, 87.9,
    86.4, 85.1, 84.7, 83.2, 82.9, 80.4, 79.8,
    78.3, 76.9, 75.4, 72.1, 70.8, 68.3,
    74.5, 65.2, 63.8, 61.4, 58.9, 54.3,
    52.1, 51.7, 49.4, 45.2, 41.8, 39.6,
    38.2, 36.9, 34.5, 30.1
  )
)
```

```{r}
#| label: fig-ranking
#| fig-cap: "Penta3 coverage ranges from 95% in Anambra to 30% in Kebbi, with a clear divide between southern and northern states."
#| fig-height: 9
#| fig-width: 8

penta3_state |>
  mutate(state = fct_reorder(state, coverage)) |>
  ggplot(aes(x = coverage, y = state)) +
  geom_col(fill = "#2C7BB6", width = 0.7) +
  geom_text(
    aes(label = paste0(coverage, "%")),
    hjust = -0.15,
    size = 3,
    colour = "#333333"
  ) +
  geom_vline(xintercept = 80, linetype = "dashed",
             colour = "#E84855", linewidth = 0.5) +
  annotate("text", x = 80.5, y = 2, label = "80% target",
           hjust = 0, size = 3, colour = "#E84855") +
  scale_x_continuous(
    limits = c(0, 105),
    labels = label_percent(scale = 1)
  ) +
  labs(
    title = "Only 14 of 36 states meet the 80% Penta3 coverage target",
    subtitle = "Estimated Penta3 coverage by state, Nigeria DHS 2023-24",
    x = "Penta3 coverage (%)",
    y = NULL,
    caption = "Source: Nigeria DHS 2023-24 (PR157). Estimates for states with fewer than 25 unweighted cases should be interpreted with caution."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(colour = "#555555", size = 10),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 9)
  )
```

Notice what the ordered horizontal bar chart does immediately: the gap between the top and bottom states is visible in a single glance. A policymaker does not need to read every label. The chart does the ranking for them.

::: {.callout-tip}
**Design decision:** The dashed reference line at 80% adds a policy benchmark without requiring a separate annotation block. A reader can instantly see which states meet the target and which do not. Whenever you have a national target or threshold, put it on the chart.
:::

------------------------------------------------------------------------

## Comparison type 2: Part-to-whole

**The question:** What share of the total does each group represent?

**Chart type to use:** Stacked bar chart (for one category) or 100% stacked bar chart (for comparing shares across groups).

**Why not a pie chart?** Pie charts require the reader to judge angles. Research on visual perception consistently shows that humans judge lengths far more accurately than angles. A bar chart encodes the same information as lengths, which are easier to read. Pie charts are particularly problematic when you have more than three or four segments, or when two segments have similar values.

The one place a pie chart works is when you have two segments (yes/no, vaccinated/unvaccinated) and you want to highlight a dominant majority. Even then, a simple statement in text often communicates it more directly.

### Example: Vaccine delivery channel by zone, Nigeria DHS 2023--24

```{r}
#| label: part-whole-data

delivery_channel <- data.frame(
  zone = rep(c("North Central", "North East", "North West",
               "South East", "South South", "South West"), each = 3),
  channel = rep(c("Fixed facility", "Outreach", "Other/unknown"), times = 6),
  percent = c(
    62, 28, 10,
    41, 47, 12,
    38, 51, 11,
    78, 15,  7,
    71, 22,  7,
    80, 14,  6
  )
)
```

```{r}
#| label: fig-part-whole
#| fig-cap: "Northern zones rely more heavily on outreach delivery than southern zones, where fixed facility vaccination is dominant."
#| fig-height: 5
#| fig-width: 8

zone_order <- delivery_channel |>
  filter(channel == "Fixed facility") |>
  arrange(percent) |>
  pull(zone)

delivery_channel |>
  mutate(
    zone    = factor(zone, levels = zone_order),
    channel = factor(channel, levels = c("Fixed facility", "Outreach", "Other/unknown"))
  ) |>
  ggplot(aes(x = percent, y = zone, fill = channel)) +
  geom_col(position = "fill", width = 0.65) +
  scale_x_continuous(labels = label_percent()) +
  scale_fill_manual(
    values = c(
      "Fixed facility" = "#2C7BB6",
      "Outreach"       = "#ABD9E9",
      "Other/unknown"  = "#D9D9D9"
    )
  ) +
  labs(
    title = "Outreach delivery drives vaccination in northern zones",
    subtitle = "Share of Penta3 doses by delivery channel and zone, Nigeria DHS 2023-24",
    x = NULL,
    y = NULL,
    fill = "Delivery channel",
    caption = "Source: Nigeria DHS 2023-24 (PR157). Illustrative values based on DHS methodology."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(colour = "#555555", size = 10),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

::: {.callout-tip}
**Design decision:** The 100% stacked bar (using `position = "fill"`) normalises each row to 100%, making the proportional comparison across zones the focus. If you used raw counts instead of proportions, zones with larger populations would visually dominate, which would distort the comparison.
:::

------------------------------------------------------------------------

## Comparison type 3: Change over time

**The question:** How has this indicator moved across time?

**Chart type to use:** Line chart.

**Why not a bar chart for time series?** Bar charts treat each time point as a discrete category. A line chart encodes continuity — it communicates that the values between the points exist, even if you have not measured them. For time series data, continuity is usually the right mental model. Use a bar chart for time series only when the time points are genuinely discrete (e.g. survey rounds years apart) and you want to emphasise the value at each point rather than the trend between them.

### Example: Penta3 coverage trend across DHS survey rounds, Nigeria

```{r}
#| label: time-series-data

penta3_trend <- data.frame(
  year  = c(2003, 2008, 2013, 2018, 2023),
  national = c(22.4, 41.5, 48.9, 57.3, 64.7),
  north    = c(12.1, 26.3, 33.8, 42.1, 51.4),
  south    = c(35.6, 58.4, 67.2, 75.8, 80.3)
)

penta3_long <- penta3_trend |>
  pivot_longer(-year, names_to = "group", values_to = "coverage") |>
  mutate(group = case_when(
    group == "national" ~ "National",
    group == "north"    ~ "North (combined)",
    group == "south"    ~ "South (combined)"
  ))
```

```{r}
#| label: fig-time-series
#| fig-cap: "National Penta3 coverage has tripled since 2003, but a persistent north-south gap remains across all survey rounds."
#| fig-height: 5
#| fig-width: 8

ggplot(penta3_long, aes(x = year, y = coverage, colour = group, group = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 3) +
  geom_text(
    data = penta3_long |> filter(year == 2023),
    aes(label = paste0(group, "\n", coverage, "%")),
    hjust = -0.1,
    size = 3,
    lineheight = 0.9
  ) +
  scale_x_continuous(
    breaks = c(2003, 2008, 2013, 2018, 2023),
    limits = c(2003, 2029)
  ) +
  scale_y_continuous(
    limits = c(0, 100),
    labels = label_percent(scale = 1)
  ) +
  scale_colour_manual(
    values = c(
      "National"         = "#2C7BB6",
      "North (combined)" = "#E84855",
      "South (combined)" = "#43AA8B"
    )
  ) +
  labs(
    title = "Coverage has grown nationally, but the north-south gap has not closed",
    subtitle = "Penta3 coverage across DHS survey rounds, Nigeria 2003-2023",
    x = "DHS survey year",
    y = "Penta3 coverage (%)",
    caption = "Source: Nigeria DHS series. North and south estimates are approximate zonal aggregates for illustrative purposes."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(colour = "#555555", size = 10),
    legend.position = "none",
    panel.grid.minor = element_blank()
  )
```

::: {.callout-tip}
**Design decision:** Direct labels replace the legend. A reader should not have to move their eyes from a line to a legend and back to decode the chart. Placing the label at the end of each line keeps the reading path clean. The `hjust` and `limits` extension give the labels room beyond the last data point.
:::

------------------------------------------------------------------------

## Comparison type 4: Distribution

**The question:** How spread out are the values across states or LGAs? Where are the extremes?

**Chart type to use:** Dot plot or box plot.

This is one of the most underused chart types in public health data work. Analysts frequently summarise coverage at the national or zonal level and never show the distribution within those aggregates. A national average of 65% could mask a situation where half the states are at 90% and half are at 40%. The distribution chart reveals this.

### Example: Distribution of Penta3 coverage across states by zone

```{r}
#| label: distribution-data

penta3_zone <- data.frame(
  state = c(
    "Benue", "FCT Abuja", "Kogi", "Kwara", "Nasarawa", "Niger", "Plateau",
    "Adamawa", "Bauchi", "Borno", "Gombe", "Taraba", "Yobe",
    "Jigawa", "Kaduna", "Kano", "Katsina", "Kebbi", "Sokoto", "Zamfara",
    "Abia", "Anambra", "Ebonyi", "Enugu", "Imo",
    "Akwa Ibom", "Bayelsa", "Cross River", "Delta", "Edo", "Rivers",
    "Ekiti", "Lagos", "Ogun", "Ondo", "Osun", "Oyo"
  ),
  zone = c(
    rep("North Central", 7),
    rep("North East", 6),
    rep("North West", 7),
    rep("South East", 5),
    rep("South South", 6),
    rep("South West", 6)
  ),
  coverage = c(
    75.4, 74.5, 61.4, 63.8, 72.1, 65.2, 70.8,
    34.5, 38.2, 39.6, 36.9, 68.3, 41.8,
    49.4, 58.9, 51.7, 52.1, 30.1, 54.3, 45.2,
    87.9, 95.4, 78.3, 84.7, 89.8,
    76.9, 71.2, 80.4, 79.8, 83.2, 86.4,
    90.3, 91.6, 88.2, 82.9, 89.5, 85.1
  )
)
```

```{r}
#| label: fig-distribution
#| fig-cap: "The North West shows the widest variation in coverage across its states, while the South East is more consistently high."
#| fig-height: 5.5
#| fig-width: 8

penta3_zone |>
  mutate(zone = fct_reorder(zone, coverage, .fun = median)) |>
  ggplot(aes(x = coverage, y = zone)) +
  geom_boxplot(
    fill = "#ABD9E9", colour = "#2C7BB6",
    outlier.shape = NA, width = 0.5
  ) +
  geom_jitter(
    aes(colour = coverage < 50),
    height = 0.15, width = 0,
    size = 2.5, alpha = 0.8
  ) +
  scale_colour_manual(
    values = c("FALSE" = "#2C7BB6", "TRUE" = "#E84855"),
    guide = "none"
  ) +
  scale_x_continuous(
    limits = c(20, 100),
    labels = label_percent(scale = 1)
  ) +
  geom_vline(xintercept = 80, linetype = "dashed",
             colour = "#E84855", linewidth = 0.5) +
  labs(
    title = "Wide within-zone variation in the North masks pockets of very low coverage",
    subtitle = "Distribution of Penta3 coverage by state within zone, Nigeria DHS 2023-24",
    x = "Penta3 coverage (%)",
    y = NULL,
    caption = "Source: Nigeria DHS 2023-24 (PR157). Red points indicate states below 50% coverage."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(colour = "#555555", size = 10),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

::: {.callout-tip}
**Design decision:** The box plot shows the median and interquartile range for each zone. Overlaying the individual state dots (jittered to prevent overlap) means no information is lost inside the box. Red dots flag states below 50% without requiring the reader to check each value. This is dual encoding: colour and position together carry the message.
:::

------------------------------------------------------------------------

## Comparison type 5: Relationship

**The question:** Does one variable move with another? Is there a pattern between two indicators?

**Chart type to use:** Scatter plot.

In public health analysis, scatter plots are used less frequently than they should be. Common questions this comparison type answers: do states with lower Penta1 coverage also have worse dropout? Does household wealth predict full vaccination? Is there a relationship between birth order and immunisation status? These questions require a scatter plot.

### Example: Penta1 coverage versus Penta1-to-Penta3 dropout rate by state

```{r}
#| label: relationship-data

penta_scatter <- data.frame(
  state = c(
    "Anambra", "Lagos", "Ekiti", "Imo", "Osun", "Ogun", "Abia",
    "Rivers", "Oyo", "Enugu", "Edo", "Ondo", "Cross River", "Delta",
    "Ebonyi", "Akwa Ibom", "Benue", "Nasarawa", "Plateau", "Taraba",
    "FCT Abuja", "Niger", "Kwara", "Kogi", "Kaduna", "Sokoto",
    "Katsina", "Kano", "Jigawa", "Zamfara", "Yobe", "Borno",
    "Bauchi", "Gombe", "Adamawa", "Kebbi"
  ),
  penta1 = c(
    97.1, 93.8, 92.4, 91.5, 92.0, 90.5, 90.1,
    89.4, 88.6, 87.2, 86.0, 85.7, 83.9, 83.2,
    82.4, 80.6, 79.8, 76.3, 75.1, 74.6,
    78.2, 71.5, 70.1, 67.8, 66.4, 63.2,
    61.8, 61.2, 58.7, 55.4, 51.9, 49.8,
    47.6, 46.1, 43.8, 39.2
  ),
  dropout = c(
    1.7, 2.2, 2.1, 1.7, 2.5, 2.3, 2.2,
    3.0, 3.5, 2.5, 2.8, 2.8, 3.5, 3.4,
    4.1, 3.7, 4.4, 4.2, 4.3, 6.3,
    3.7, 6.3, 6.3, 6.4, 7.5, 8.9,
    9.7, 9.5, 9.3, 10.2, 10.1, 10.2,
    9.4, 9.2, 9.3, 9.1
  ),
  zone = c(
    rep("South", 16),
    rep("North Central", 5),
    rep("North", 15)
  )
)
```

```{r}
#| label: fig-relationship
#| fig-cap: "States with lower Penta1 coverage also tend to have higher dropout rates, suggesting that access and retention challenges often coincide."
#| fig-height: 5.5
#| fig-width: 8

ggplot(penta_scatter, aes(x = penta1, y = dropout, colour = zone)) +
  geom_point(size = 3, alpha = 0.85) +
  geom_smooth(method = "lm", se = FALSE,
              colour = "#555555", linewidth = 0.7, linetype = "dashed") +
  ggrepel::geom_text_repel(
    data = penta_scatter |>
      filter(dropout > 9 | dropout < 2.5 | penta1 > 95),
    aes(label = state),
    size = 2.8,
    max.overlaps = 10
  ) +
  scale_x_continuous(
    limits = c(35, 100),
    labels = label_percent(scale = 1)
  ) +
  scale_y_continuous(
    limits = c(0, 13),
    labels = function(x) paste0(x, "%")
  ) +
  scale_colour_manual(
    values = c(
      "South"         = "#43AA8B",
      "North Central" = "#F9C74F",
      "North"         = "#E84855"
    )
  ) +
  labs(
    title = "States with the lowest access also have the worst retention",
    subtitle = "Penta1 coverage vs Penta1-to-Penta3 dropout rate by state, Nigeria DHS 2023-24",
    x = "Penta1 coverage (%)",
    y = "Penta1-to-Penta3 dropout rate (%)",
    colour = "Region",
    caption = "Source: Nigeria DHS 2023-24 (PR157). Dropout estimated from Penta1 and Penta3 state-level figures."
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", size = 12),
    plot.subtitle = element_text(colour = "#555555", size = 10),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

::: {.callout-note}
This chart uses the `ggrepel` package for non-overlapping labels. Install it with `install.packages("ggrepel")` and load it with `library(ggrepel)` at the top of your script.
:::

::: {.callout-tip}
**Design decision:** Not all state labels are shown. Labelling every point creates clutter and makes the chart harder to read. Here, only the outliers are labelled: states with the highest dropout, lowest coverage, and best-performing states. These are the points a policymaker is most likely to ask about, so they get names.
:::

------------------------------------------------------------------------

## The chart type decision framework

Use this table as a quick reference when you are deciding which chart to build.

| You want to show... | Use this chart | Avoid |
|---|---|---|
| Which group is highest or lowest | Horizontal bar (ordered) | Pie chart, unsorted bars |
| Shares of a total | 100% stacked bar | Pie chart (more than 3 segments) |
| How a value has changed over time | Line chart | Bar chart for continuous series |
| How values are spread within a group | Box plot, dot plot | Bar chart showing only the mean |
| Whether two variables move together | Scatter plot | Dual-axis line chart |

The most common mismatch in public health reporting is using a bar chart to show change over time when a line chart would make the trend far clearer. The second most common is using a national or zonal average bar chart when a distribution chart would show the range of performance that the average conceals.

------------------------------------------------------------------------

## What to watch out for

**Pie charts with more than three segments.** The standard recommendation in data visualisation practice is to use a pie chart only when you have two or three segments and one segment is dominant. Beyond three segments, bar charts almost always communicate the proportions more clearly.

**Dual-axis charts.** A chart with two y-axes, each on a different scale, asks the reader to hold two scales in mind simultaneously. The visual relationship between the two lines appears to change depending on how the scales are set. For most policy audiences, two separate charts are easier to interpret than one dual-axis chart.

**3D charts.** Three-dimensional bar and pie charts distort the visual proportions of the data. The bars or slices at the front of the chart appear larger than those at the back, even when the underlying values are identical. This is not a style preference. It is a perceptual distortion that introduces inaccuracy into how the data is read.

**Smoothed trend lines without uncertainty.** A smooth line through scattered data points implies more precision than the underlying data supports. When showing trends across a small number of data points (such as DHS survey rounds five years apart), showing the individual points rather than a fitted curve is usually more honest.

------------------------------------------------------------------------

## Summary

Chart type choice is not a cosmetic decision. The chart type you select determines what comparisons your audience can make, and therefore what conclusions they are able to draw.

The framework in this tutorial is simple: identify the comparison type first, then select the chart that encodes it most directly. Ranking needs ordered horizontal bars. Change over time needs lines. Distribution needs dot plots or box plots. Proportions need stacked bars. Relationships need scatter plots.

The next tutorial in this series covers small multiples and faceted charts in R: how to show the same comparison across multiple groups without building a separate chart for each one.

------------------------------------------------------------------------

## Further reading

-   Schwabish, J. (2021). *Better Data Visualizations: A Guide for Scholars, Researchers, and Wonks*. Columbia University Press.
-   Wilke, C.O. (2019). *Fundamentals of Data Visualization*. O'Reilly. Available free at <https://clauswilke.com/dataviz>
-   Cairo, A. (2016). *The Truthful Art: Data, Charts, and Maps for Communication*. New Riders.
